A general and efficient representation of Ancestral Recombination Graphs

### 2022-04-08 paper.tex
More minor tweaks
Fixes #209 and a few readability fixes

\node [anchor=north west] at (-14,6) {A};
\node [nodelabel,anchor=north west] at ($(-13,5)$) {
\begin{tabular}{c|c}
% \multicolumn{2}{c}{Breakpoint}\\
node & breakpoint\\
\hline
$\textsf{a}$ & $\varnothing$ \\
$\textsf{b}$ & $\varnothing$ \\
$\textsf{c}$ & $\varnothing$ \\
$\textsf{d}$ & $x$ \\
$\textsf{e}$ & $\varnothing$ \\
$\textsf{f}$ & $\varnothing$ \\
$\textsf{g}$ & $\varnothing$ \\
\end{tabular}};

@@

\node [anchor=north west] at (9,6) {C};
\node [nodelabel,anchor=north west] at ($(10,5)$) {
\begin{tabular}{ll}
Edges\\
\hline
$(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
$(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
$(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
$(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
$(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
$(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
$(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
\end{tabular}};

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A classical ARG (subfigure B), contrasting node with edge breakpoint annotations.
(A) recombination breakpoint information encoded on \emph{nodes}.
This encoding is limited in the patterns of inheritance that can be
described and requires that edges be ordered to avoid ambiguity.
(C) breakpoint annotations associated with \emph{edges}.
Associating the interval of genome inherited by the child node
from the parent with the corresponding edge allows
any pattern genetic inheritance to be described and does
not require edges to be ordered.
}
\end{figure}

% What do we mean by an eARG, formally?
Formally, we can define an eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise. (This formulation is equivalent to stipulating that
a recombination node is of a different ``type'' to other nodes,
with breakpoints only associated with this specific node type).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

@@

% it's not reasonable to assume that one type of event happens
% at a time any more.
The eARG is also limited in the types of ancestral history that can
be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough that,
with high probability, the number of concurrent lineages in the ARG remains
much smaller than $N$ at each time. While these approximations are
reasonable to make for small datasets, such as
the classic \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster}
ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
as used in the ``\nameref{ARG_inference_methods}'' section),
they are highly questionable in modern datasets. % absurd really
For example, many
human datasets consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
% TODO make this better. Ask Gregor for some refs?
% Basically, Ag datasets make the assumption of isolated
% rare recombinations absurd.
Agricultural datasets such as the 1000 bull genomes~\citep{hayes20191000}
combine genome sequence data with detailed pedigrees, with
sample data from multiple generations.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust to such
basic violations to the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, and we can be
certain that this is also true in the incredibly rich
datasets that are available today.

>>>>

\node [anchor=north west] at (-14,6) {A};
\node [nodelabel,anchor=north west] at ($(-13,5)$) {
\begin{tabular}{c|c}
% \multicolumn{2}{c}{Breakpoints}\\
Nodes & Breakpoints\\
\hline
$\textsf{a}$ & $\varnothing$ \\
$\textsf{b}$ & $\varnothing$ \\
$\textsf{c}$ & $\varnothing$ \\
$\textsf{d}$ & $x$ \\
$\textsf{e}$ & $\varnothing$ \\
$\textsf{f}$ & $\varnothing$ \\
$\textsf{g}$ & $\varnothing$ \\
\end{tabular}};

@@

\node [anchor=north west] at (9,6) {C};
\node [nodelabel,anchor=north west] at ($(10,5)$) {
\begin{tabular}{ll}
Edges & Intervals\\
\hline
$(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
$(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
$(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
$(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
$(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
$(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
$(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
\end{tabular}};

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A classical ARG (subfigure B), contrasting node with edge breakpoint annotations.
(A) recombination breakpoint information encoded on \emph{nodes}.
This encoding is limited in the patterns of inheritance that can be
described and requires that edges be ordered to avoid ambiguity.
(C) breakpoint annotations associated with \emph{edges}.
Associating the interval of genome inherited by the child node
from the parent with the corresponding edge allows
any pattern genetic inheritance to be described and does
not require edges to be ordered. In this example, each edge is associated
with a single interval, but more generally, multiple non-overlapping intervals
can be associated with a single edge.
}
\end{figure}

% What do we mean by an eARG, formally?
Formally, we can define the classical eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise. (This formulation is equivalent to stipulating that
a recombination node is of a different ``type'' to other nodes,
with breakpoints only associated with that node type).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

@@

% it's not reasonable to assume that one type of event happens
% at a time any more.
The eARG is also limited in the types of ancestral history that can
be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough that,
with high probability, the number of concurrent lineages in the ARG remains
much smaller than $N$ at each time. These approximations are
reasonable to make for small datasets, such as
the classic \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster}
ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
see the ``\nameref{ARG_inference_methods}'' section). However,
they are highly questionable in modern datasets. % absurd really
For example, many
human datasets consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
% TODO make this better. Ask Gregor for some refs?
% Basically, Ag datasets make the assumption of isolated
% rare recombinations absurd.
Agricultural datasets such as the 1000 bull genomes~\citep{hayes20191000}
combine genome sequence data with detailed pedigrees, with
sample data from multiple generations.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust to such
basic violations to the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, and we can be
certain that this is also true in the incredibly rich
datasets that are available today.
